Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-10 cis-perfect-testing.tex
Fix chapter numbering
Bulk of ATACCC chapter (no results yet)
Various edits

\begin{document}
\section{Definitions}
\todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
\begin{itemize}
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\end{itemize}

@@

Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial, see \cref{E-distributions}).
I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite.
Using the composite simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

Simulation was performed in R~4.2.0~\autocite{R4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.

>>>>

\begin{document}
\ifSubfilesClassLoaded{

\section{Definitions}
\todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
\begin{itemize}
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\end{itemize}

\setcounter{chapter}{4}
}

@@

Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB_c(\mu, r)$.
I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Normally-distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

Simulation was performed in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.

### 2023-11-10 distributions.tex
Various edits

\section{Discrete distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pmf \\
\hline \\
Binomial & $\Bin(n, p)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
Poisson & $\Poi(\lambda)$ & $\lambda > 0$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
Negative Binomial (standard) & $\NB(r, p)$ & $r > 0$ (number of failures until stopping)\newline $0 < p < 1$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x! \Gamma(r)} p^x (1-p)^r$ \\
Negative Binomial (centred) & $\NB(\mu, r)$ & $\mu > 0$ (mean) \newline r (dispersion) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x! \Gamma(r)} \left( \frac{r}{r+\mu} \right)^\mu \left( \frac{\mu}{r+\mu} \right)^k$ \\
Beta-Binomial (centred) & $\BB(n, p, \rho)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (success probability)\newline $\rho > 0$ (overdispersion) & $np$ & $np(1-p)(1+n\rho)(1+\rho)^{-1}$ & ${n \choose x} \frac{B(x+p/\rho, n-x+(1-p)/\rho)}{B(p/\rho, (1-p)/\rho)}$ \\
Multinomial & $\MN(n, \vec{p})$ & $n \in \nats$ (number of trials, integer)\newline $\vec{p} > 0$ (vector of $k$ probabilities, $\sum_{i=1}^k p_i = 1$) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
\end{tabular}

>>>>

\section{Discrete distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pmf \\
\hline \\
Binomial & $\Bin(n, p)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
Poisson & $\Poi(\lambda)$ & $\lambda > 0$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
Negative Binomial (standard) & $\NBs(r, p)$ & $r > 0$ (number of failures until stopping)\newline $0 < p < 1$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x! \Gamma(r)} p^x (1-p)^r$ \\
Negative Binomial (centred) & $\NBc(\mu, r)$ & $\mu > 0$ (mean) \newline r (dispersion) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x! \Gamma(r)} \left( \frac{r}{r+\mu} \right)^\mu \left( \frac{\mu}{r+\mu} \right)^k$ \\
Beta-Binomial (centred) & $\BB(n, p, \rho)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (success probability)\newline $\rho > 0$ (overdispersion) & $np$ & $np(1-p)(1+n\rho)(1+\rho)^{-1}$ & ${n \choose x} \frac{B(x+p/\rho, n-x+(1-p)/\rho)}{B(p/\rho, (1-p)/\rho)}$ \\
Multinomial & $\MN(n, \vec{p})$ & $n \in \nats$ (number of trials, integer)\newline $\vec{p} > 0$ (vector of $k$ probabilities, $\sum_{i=1}^k p_i = 1$) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
\end{tabular}
